Full Brain Blood-Oxygen-Level-Dependent Signal Parameter Estimation Using Particle Filters

### 2010-09-02 doc/paper/main.tex
small fixes

These shortcomings have led to several other separate attempts. In particular 
Riera et al. linearized the noise and thus performed regression between
the model and steps of the BOLD output \cite{Riera2003}. This aproach
had the benefit of a Jacobian,  but at the same time removes all DC 
signal. For this reason, the type of activation limits the
algorithms' ability to learn the model.
In Vakorin et al., a combination of Genetic Algorithms and Simulated 
Annealing were used to estimate not only the parameters, but the 
true stimuli driving the BOLD signal \cite{Vakorin2007}. 
This addresses the inherent uncertainty of exactly where and when 
stimuli actually get applied. Unfortunately this algorithm was extremely
slow; taking hours or days per single time course.
In Johnston et al. alternating estimates of $P(X_t | \Theta, Y_t)$ and 
$P(\Theta | Y_t)$ were calculated, to maximize $P(X_t, \Theta | Y_t)$. 
$P(X_t | \Theta, Y_t)$ was calculated using a particle filter 
\cite{Johnston2008}. This however was quite slow, and the results
were inconsistent with other similar works. 
Murray used a particle filter, similar to \cite{Johnston2008}, but
held the model parameters constant; thus all error in the output
was treated as the result of noise in the underlying parameters 
\cite{Murray2008}. Murray's results showed that differences in
BOLD output cannot be well explained by error in the state variables.
Hu et al. used an unscented kalman filter (UKF) to estimate parameters,
and filtered parameter sets inconsistent with the output  \cite{Hu2009}.
 Furthermore, Birn et al. and Yacoub et al. showed that many
of the characteristics of the BOLD output depend on the region 
of interest \cite{Birn2001, Yacoub2006}. For this reason, a single
Hemodynamic Response Function is not likely to accurately estimate
activation across the full brain.

@@

Although Buxton et al. demonstrated that a well chosen flow waveform could 
explain most features of the BOLD signal, it stopped short of proposing a
realistic waveform for the CBF and CMRO2 \cite{Buxton1998}. Friston et al. 
gave a reasonable and simple expression for CBF input based on a flow 
inducing signal and in the same work proposed a simple method
of estimating metabolic rate: as a linear function of the inward 
blood flow \cite{Friston2000}. By combining these equations with 
the balloon model from Buxton et al., it is possible to predict 
the BOLD signal directly from a stimulus time course:
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s\\
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\label{eq:bold}
\end{eqnarray}
where $s$ is a flow inducing signal, $f$ is the input blood flow (CBF),
$v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin. The parameters controlling blood flow are 
$\epsilon$, which is a neuronal efficiency term, $u(t)$ which is 
the stimulus, and $\tau_f$, $\tau_s$ which are time constants. 
The parameters for the evolution of blood volume are $E_0$ which 
the resting metabolic rate and $\alpha$ which is Grubb's parameter 
controlling the balloon model.  $\tau_0$ is a single time constant 
controlling the speed of $v$ and $q$.

@@

\subsection{Particle Filter}
\label{sec:ParticleFilters}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a time series of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. Unlike the UFK, sample points are kept
and continuously propagated forward, rather than being used to calculate
the first two moments. Thus, the posterior distribution is
stored as an empircal distribution made up of sample points.
Particle filters are preferred when the model is nonlinear, 
and non-Gaussian. This section based primarily on 
Arulampalam et al. and Thrun et al. \cite{Arulampalam2002a, Thrun2005}.

@@

\subsubsection{Sequential Importance Resampling}
\label{sec:ParticleFilterResampling}
As a consequence of the wide prior distribution 
(required for a proper discretization of a continuous
distribution), a significant proportion of particles quickly
develop insignificant weights. 
Ideally the computation time spent on tails would be proportional to 
the actual size of the tails.  The case where a large number of the 
weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal measurement of degeneracy based on the particles' true 
weight is given \cite{Liu1998b}. An approximation of 
the effective number of particles is also given: 
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
To alleviate particle degeneracy, a technique known as resampling 
may be applied. The idea of resampling is to 
draw from the approximate posterior, thus generating 
a replica of the posterior probability with 
a matching support. Each particle in the new distribution,
$\hat{x}_j$, is drawn from the old distribution 
(see Algorithm \autoref{alg:Resampling}):
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

@@

Because of the ill-defined nature of the parameters, the results
of Friston et al. may have been wider than true parameter distribution
\cite{Friston2002}. Future works that attempt to calculate parameters
of the BOLD model should be run multiple times to ensure consistency,
which most likely cannot be attained. This result is particularly
troublesome given the number of papers that use those parameters for
a prior distribution.

>>>>

These shortcomings have led to several other separate attempts. In particular 
Riera et al. linearized the noise and thus performed regression between
the model and steps of the BOLD output \cite{Riera2003}. This aproach
had the benefit of a Jacobian,  but at the same time removes all DC 
signal. For this reason, the type of activation limits the
algorithms' ability to learn the model.
In Vakorin et al., a combination of Genetic Algorithms and Simulated 
Annealing were used to estimate not only the parameters, but the 
true stimuli driving the BOLD signal \cite{Vakorin2007}. 
This addresses the inherent uncertainty of exactly where and when 
stimuli actually get applied. Unfortunately this algorithm was extremely
slow; taking hours or days per single time course.
In Johnston et al. alternating estimates of $P(X_t | \Theta, Y_t)$ and 
$P(\Theta | Y_t)$ were calculated, to maximize $P(X_t, \Theta | Y_t)$. 
$P(X_t | \Theta, Y_t)$ was calculated using a particle filter 
\cite{Johnston2007}. This however was quite slow, and the results
were inconsistent with other similar works. 
Murray used a particle filter, similar to \cite{Johnston2007}, but
held the model parameters constant; thus all error in the output
was treated as the result of noise in the underlying parameters 
\cite{Murray2008}. Murray's results showed that differences in
BOLD output cannot be well explained by error in the state variables.
Hu et al. used an unscented kalman filter (UKF) to estimate parameters,
and filtered parameter sets inconsistent with the output  \cite{Hu2009}.
 Furthermore, Birn et al. and Yacoub et al. showed that many
of the characteristics of the BOLD output depend on the region 
of interest \cite{Birn2001, Yacoub2006}. For this reason, a single
Hemodynamic Response Function is not likely to accurately estimate
activation across the full brain.

@@

Although Buxton et al. demonstrated that a well chosen flow waveform could 
explain most features of the BOLD signal, it stopped short of proposing a
realistic waveform for the CBF and CMRO2 \cite{Buxton1998}. Friston et al. \cite{Friston2000}
gave a reasonable and simple expression for CBF input based on a flow 
inducing signal and in the same work proposed a simple method
of estimating metabolic rate: as a linear function of the inward 
blood flow. By combining these equations with 
the balloon model from Buxton et al., it is possible to predict 
the BOLD signal directly from a stimulus time course:
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s\\
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\label{eq:bold}
\end{eqnarray}
where $s$ is a flow inducing signal, $f$ is the input blood flow (CBF),
$v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin. The parameters controlling blood flow are 
$\epsilon$, which is a neuronal efficiency term, $u(t)$ which is 
the stimulus, and $\tau_f$, $\tau_s$ which are time constants. 
The parameters for the evolution of blood volume are $E_0$ which 
the resting metabolic rate and $\alpha$ which is Grubb's parameter 
controlling the balloon model.  $\tau_0$ is a single time constant 
controlling the speed of $v$ and $q$.

@@

\subsection{Particle Filter}
\label{sec:ParticleFilters}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a time series of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. Unlike the UFK, sample points are kept
and continuously propagated forward, rather than being used to calculate
the first two moments. Thus, the posterior distribution is
stored as an empircal distribution made up of sample points.
Particle filters are preferred when the model is nonlinear, 
and non-Gaussian. This section based primarily on 
Arulampalam et al. \cite{Arulampalam2002a} and Thrun et al. \cite{thrun2008probabilistic}.

@@

\subsubsection{Sequential Importance Resampling}
\label{sec:ParticleFilterResampling}
As a consequence of the wide prior distribution 
(required for a proper discretization of a continuous
distribution), a significant proportion of particles quickly
develop insignificant weights. 
Ideally the computation time spent on tails would be proportional to 
the actual size of the tails.  The case where a large number of the 
weights have become extremely small
is called particle degeneracy. In  Lui et al. 
an ideal measurement of degeneracy based on the particles' true 
weight is given \cite{Liu1998}. An approximation of 
the effective number of particles is also given: 
\begin{equation}
N_{eff} \approx \frac{\sum_{i=0}^{N_p} w_i}{\sum_{i=0}^{N_p} w_i^2}
\label{eq:neff}
\end{equation}
To alleviate particle degeneracy, a technique known as resampling 
may be applied. The idea of resampling is to 
draw from the approximate posterior, thus generating 
a replica of the posterior probability with 
a matching support. Each particle in the new distribution,
$\hat{x}_j$, is drawn from the old distribution 
(see Algorithm \autoref{alg:Resampling}):
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=0}^{N_p} w^i_k\delta(x - x^i_k)\right)
\end{equation}

@@

Because of the ill-defined nature of the parameters, the results
of Friston et al. may have been wider than true parameter distribution
\cite{Friston2002b}. Future works that attempt to calculate parameters
of the BOLD model should be run multiple times to ensure consistency,
which most likely cannot be attained. This result is particularly
troublesome given the number of papers that use those parameters for
a prior distribution.
